\textcolor{red}{The Geometry of Local Spectral Clustering}

### 2021-01-03 jmlr/revision/local_spectral_clustering_jmlr_revision_2.tex
Completed fixed graph proofs.
Added proofs of convergence of random graph functionals.

\subsection{Review: worst-case guarantees for local spectral clustering}
As mentioned, to date most analysis of local clustering has focused on worst-case guarantees, defined with respect to functionals of an a priori fixed graph $G = (V,E)$.  For instance, \cite{andersen2006} analyze the normalized cut of the cluster estimate $\wh{C}$ output by PPR. For a set $C \subseteq V$ with complement $C^c = V \!\setminus\! C$, the \emph{cut} and \emph{volume} are respectively,
\begin{equation}
\label{eqn:cut_volume}
\cut(C;G) := \sum_{u \in C} \sum_{v \in C^c}
\1\{(u,v) \in E\},~~ \vol(C; G) := \sum_{u \in C}  \sum_{v \in V} \1\{(u,v) \in E\},
\end{equation}
and the \emph{normalized cut} of $C$ is
\begin{equation}
\label{eqn:normalized_cut}
\Phi(C; G) := \frac{\cut(C;G)}{\min \set{\vol(C; G), \vol(C^c; G)}}.
\end{equation} 
\cite{andersen2006} show that when PPR is appropriately seeded within a candidate cluster $C \subseteq V$, the normalized cut of the candidate cluster is upper bounded $\Phi(\wh{C};G) \lesssim \sqrt{\Phi(C;G)}$. \cite{zhu2013} build on this: they introduce a second functional, the \emph{conductance} $\Psi(C;G)$, defined as 
\begin{equation}
\label{eqn:conductance}
\Psi(C;G) := \min_{S \subseteq C} \Phi(S;G[C]),
\end{equation}
where $G[C] = (C,E \cap (C \times C))$ is the subgraph of $G$ induced by $C$---and show that if $\Phi(C;G)$ is much smaller than $\Psi(C;G)^2$, then (in addition to having a small normalized cut) the cluster estimate $\wh{C}$ approximately recovers $C$. Our own analysis will build on that of~\cite{zhu2013}, and we will give a more detailed summary of their results in Section~\ref{sec:ub_symmetric_set_difference}.
For now, we merely reiterate that the conclusions of \citep{andersen2006,zhu2013} cannot be straightforwardly applied to our statistical setting, where the input data are random samples $\{x_1,\ldots,x_n\}$ drawn from a distribution $\mathbb{P}$, the graph $G$ is a random geometric graph formed by the user, and the candidate cluster is a set $\mc{C} \subset \Rd$.

@@

It is quite natural that $\Phi_{\Pbb,r}(\mc{C})$ and $\Psi_{\Pbb,r}(\mc{C})$ should help quantify the role geometry plays in local spectral clustering. Indeed, by construction these functionals are quite obviously the population-level analogues of the empirical quantities $\Phi_{n,r}(\mc{C}[X]) := \Phi(\mc{C}[X];G_{n,r})$ and $\Psi_{n,r}(\mc{C}[X]) = \Psi(\mc{C}[X];G_{n,r})$, and as we have already mentioned, these empirical quantities in turn suffice to upper bound the volume of the symmetric set difference. For this reason, similar population level functionals are used by \citep{shi2009,schiebinger2015,garciatrillos19} in the analysis of \emph{global} spectral clustering in a statistical context. We will comment more on the relationship between these works and our own results in~\ref{subsec:related_work}.

@@

% In the context of PPR clustering on a fixed graph $G$ and candidate cluster $C$, \textcolor{red}{(Allen-Zhu 13)} provide the sharpest known bounds on the size of the symmetric set difference $\Delta(\wh{C},C)$. We build on these results in the statistical context, where the graph $G_{n,r}$ is a random neighborhood graph, formed over samples $\{x_1,\ldots,x_n\}$ drawn from a distribution $\Pbb$, and the candidate cluster $\mc{C} \subset \Rd$.

%\begin{itemize}
	%\item In Proposition~\ref{prop:sample_to_population}, we show that with high probability the empirical quantities $\Phi_{n,r}(\mc{C}[X])$ and $\Psi_{\mathbb{P},r}(\mc{C}[X])$ are close to their population-level counterparts $\Phi_{\Pbb,r}(\mc{C})$ and $\Psi_{\Pbb,r}(\mc{C})$. As a consequence, in Theorem~\ref{thm:volume_ssd_ub} we show that $\Delta(\wh{C},\mc{C}[X])$ is small when $\Phi_{\Pbb,r}(\mc{C})$ is small relative to $\Psi_{\Pbb,r}(\mc{C})$. These give the first population-level guarantees for PPR clustering in a non-parametric statistical context.
%\end{itemize}

%We specialize to the case where the candidate cluster $\mc{C} = \mc{C}_{\lambda}$ is a \emph{$\lambda$-density cluster}---that is, a connected component of the upper level set $\{x: f(x) \geq \lambda\}$ (an object of central interest in the statistical clustering literature, dating back to the work of \citet{hartigan1981}). 
%\begin{itemize}
%	\item In Propositions~\ref{prop:density_cluster_normalized_cut} and~\ref{prop:density_cluster_conductance}, we derive specific upper bounds on the normalized cut $\Phi_{\Pbb,r}(\mc{C}_{\lambda})$ and conductance $\Psi_{\Pbb,r}(\mc{C}_{\lambda})$; these upper bounds depend on $\lambda$ as well as some other natural parameters. These immediately result in an upper bound on $\Delta(\wh{C},\mc{C}[X])$ in the density cluster setting (Corollary~\ref{cor:density_cluster_volume_ssd_ub}),  giving new connections between (local) spectral and density clustering.
%	\item We introduce a notion of \emph{density cluster consistency}. In Theorem~\ref{thm:density_cluster_consistent_recovery}, we give conditions under which $\wh{C}$ is a consistent estimate of the density cluster $\mc{C}_{\lambda}$.
%	\item In Theorem~\ref{thm:ppr_density_cluster_lb} we give a lower bound: we exhibit a hard class of densities $\mc{F}$ and corresponding density clusters $\mc{C}_{\lambda}$ for which the symmetric set difference $\Delta(\wh{C},\mc{C}_{\lambda}[X])$ will be provably large.
% \end{itemize}

@@

\subsection{PPR cluster recovery: the fixed graph case.}

When PPR is run on a fixed graph $G = (V,E)$ with the goal of recovering a candidate cluster $C \subset V$, \cite{zhu2013} provide the sharpest known bounds on the volume of the symmetric set difference between the cluster estimate $\wh{C}$ and candidate cluster $C$. Since these results will play a major part in our analysis, in Lemma~\ref{lem:zhu} we restate them for the convenience of the reader.\footnote{Technically speaking, Lemma~3.4 of~\cite{zhu2013} differs from Lemma~\ref{lem:zhu} by some constant factors, and for completeness we prove Lemma~\ref{lem:zhu} in our appendix. Nevertheless, to be clear the essential idea of Lemma~\ref{lem:zhu} is no different than that of~\cite{zhu2013}, and we do not claim any novelty.}

In their most general form, the results of~\citet{zhu2013} depend on the mixing time of a lazy random walk over the induced subgraph $G[C] = (C, E \cap (C \times C))$. Letting 
$W_C$ be the lazy random walk matrix over $G[C]$, we write  
\begin{equation*}
q_{v}^{(t)}(u) = e_vW_C^t e_u
\end{equation*}
for the probability that after $t$-steps, the lazy random walk originating at $v$ arrives at $u$. Assuming $G[C]$ is a connected graph, the random walk with transition probabilities given by $W_C$ is an irreducible and aperiodic Markov Chain; thus a unique stationary distribution exists and is given by \smash{$\pi_C(u) := \deg(u;G[C])/\vol(C;
	G[C])$}, where we write $\deg(u;G[C]) = \sum_{w \in C} \1\{(u,w) \in E_C\}$ for
the degree of $u$ in $G[C]$. The \emph{mixing time} over $G[C]$ is then
\begin{equation}
\label{eqn:mixing_time}
\tau_{\infty}(C;G) := \min\set{ t: \frac{\pi_C(u) - q_{v}^{(t)}(u)}
	{\pi_C(u)} \leq \frac{1}{4}, \; \text{for all $u,v \in C$}}
\end{equation}
\begin{lemma}[Lemma~3.4 of \cite{zhu2013}]
	\label{lem:zhu}
	For a set $C \subseteq V$, suppose that
	\begin{equation}
	\label{eqn:zhu_condition}
	\alpha \leq \min\Bigl\{\frac{1}{45}, \frac{1}{2\tau_{\infty}(C;G)}\Bigr\},~~ \beta \leq \frac{1}{5\vol(C;G)}
	\end{equation}
	Then there exists a set $C^g \subset C$ with $\vol(C^g;G) \geq \frac{1}{2}\vol(C^g;G)$ such that for any $v \in C^g$, the sweep cut $S_{\beta,v}$ satisfies
	\begin{equation}
	\label{eqn:zhu_ub}
	\vol(S_{\beta,v} \vartriangle C;G) \leq 6\frac{\Phi(C;G)}{\alpha \beta}.
	\end{equation}
\end{lemma}
The upper bound in~\eqref{eqn:zhu_ub} does not obviously depend on the conductance $\Psi(C;G)$. However, as \cite{zhu2013} point out, letting $\pi_{\min}(C;G) := \min_{u \in C}\{\pi_C(u)\}$ denote the minimum stationary probability over $C$, it follows from Cheeger's inequality~\citep{chung1997} that 
\begin{equation}
\label{eqn:mixing_time_cheeger}
\tau_{\infty}(C;G) \leq \frac{\log(1/\pi_{\min}(C;G))}{\Psi(C;G)^2}.
\end{equation}
Therefore, setting (for instance) $\alpha = \frac{\Psi(C;G)^2}{2\log(1/\pi_{\min}(C;G))}$ and $\wh{C} = S_{\beta_0,v}$ for $\beta_0 = \frac{1}{5 \vol(C;G)}$, we obtain from~\eqref{eqn:zhu_ub} that 
\begin{equation}
\label{eqn:zhu_ub2}
\frac{\vol(C \vartriangle \wh{C};G)}{\vol(C; G)} \leq 60\frac{\Phi(C;G) \log\bigl( 1/\pi_{\min}(G[C])\bigr)}{\Psi(C;G)^2}.
\end{equation}

\paragraph{Improved bounds on mixing time.} Having reviewed the conclusions of~\cite{zhu2013}, we return now to our own setting, where the data is not a fixed graph $G$ but instead random samples $\{x_1,\ldots,x_n\}$, and our goal is to recover a candidate cluster $\mc{C} \subset \Rd$. Ideally, we would like to apply~\eqref{eqn:zhu_ub2} with $C = \mc{C}[X]$ and $G = G_{n,r}$, replace $\Phi_{n,r}(\mc{C}[X])$ and $\Psi_{n,r}(\mc{C}[X])$ by $\Phi_{\Pbb,r}(\mc{C})$ and $\Phi_{\Pbb,r}(\mc{C})$ inside~\eqref{eqn:zhu_ub2}, and thereby obtain an upper bound on $\Delta(\wh{C};\mc{C}[X])$ that depends only on $\Pbb$ and $\mc{C}$. Unfortunately, however, there is a catch: when the graph $G = G_{n,r}$ and the candidate cluster $C = \mc{C}[X]$, as $n \to \infty$ the sample normalized cut $\Phi_{n,r}(\mc{C}[X])$ and conductance $\Psi_{n,r}(\mc{C}[X])$ each converge to their population-level analogues, but $\pi_{\min}(\mc{C}[X];G_{n,r}) \lesssim 1/n$. Therefore the right hand side of~\eqref{eqn:zhu_ub2} grows at a rate of at least $\log n$, rendering~\eqref{eqn:zhu_ub2} a vacuous upper bound when the number of samples is sufficiently large.\footnote{To be clear: $\pi_{\min}(G)$ is obviously at most $1/n$ for any graph $G$ with $n$ total vertices, not merely $G = G_{n,r}$. However, while in general it is certainly possible that $\Phi(C;G) \leq \frac{1}{\log(n)} \Psi(C;G)$, this will not be the case in our specific setting, where $G = G_{n,r}$ and $C = \mc{C}[X]$. Hence the need for a sharper bound.}

To fix this, we will improve the bound on mixing time given in~\eqref{eqn:mixing_time_cheeger}. While in the worst case (over the input graph $G$) the ``start penalty'' $\log(1/\pi_{\min}(C;G))$ is unavoidable, fortunately better bounds are possible when $G$ is a geometric graph such as $G_{n,r}$. Intuitively, this is because in geometric graphs, small sets are expanders---that is, if a set $R \subseteq V$ has sufficiently small volume, the normalized cut $\Phi(R;G)$ will be much larger than the conductance $\Psi(V; G)$. As a consequence, a random walk over $G$ should rapidly mix over all small sets $R$, and in our analysis of the mixing time we may therefore ``pretend'' as if the random walk were given a warm start over a larger set $S$. To delineate small sets $R$ from larger sets $S$, we introduce the \emph{local spread} functional $s(C; G)$, defined as\footnote{The prefactor $9/10$ in the definition of $s(G)$ is arbitrary. It can be altered to any constant less than $1$, and Proposition~\ref{prop:pointwise_mixing_time} will still hold after appropriate changes to the constants.}
\begin{equation*}
s(C; G) := \frac{9}{10} \cdot \min_{u \in C} \set{\deg(u; G[C])} \cdot \min_{u \in C} \set{\pi_C(u)}.
\end{equation*}
Using a careful analysis we can make the intuition of the prior paragraph rigorous, leading to an improved upper bound on mixing time in Proposition~\ref{prop:pointwise_mixing_time}. The proof of this proposition, as with all results in this paper, is deferred to the appendix.
\begin{proposition}
	\label{prop:pointwise_mixing_time}
	Assume $\min_{u \in V} \deg(u; G) \geq 10$. Then,
	\begin{equation}
	\label{eqn:pointwise_mixing_time}
	\tau_{\infty}(C; G) \leq \frac{2}{\Psi^2(C; G)} \ln \left(\frac{320}{s(C; G)}\right)\log_2 \left(\frac{14}{s(C; G)}\right)  + 3 \log_2 \left(\frac{14}{s(C; G)}\right) + 3
	\end{equation}
\end{proposition}

We emphasize that while Proposition~\ref{prop:pointwise_mixing_time} can be applied to \emph{any} graph $G$ (as long as the minimum degree is at least 10), it is particularly useful when $G$ is a geometric graph. We give a precise upper bound on $s(\mc{C}[X]; G_{n,r})$ in Proposition~\ref{prop:sample_to_population}, which does not grow with $n$, and in combination with Proposition~\ref{prop:pointwise_mixing_time} this allows us to remove the unwanted $\log n$ factor from the upper bound in~\eqref{eqn:mixing_time}.

\subsection{Sample-to-population results}
In Proposition~\ref{prop:sample_to_population}, we establish high probability bounds on the sample normalized cut, conductance, and local spread, in terms of their population-level analogues. We have already introduced the population normalized cut~\eqref{eqn:population_normalized_cut} and conductance~\eqref{eqn:population_conductance}; writing $s_{n,r}(\mc{C}[X]) := s(\mc{C}[X];G_{n,r})$ for the sample local spread, we now define the population local spread to be
\begin{equation}
\label{eqn:local_spread}
s_{\Pbb,r}(\mc{C}) := \frac{9}{10} \cdot \min_{x \in \mc{C}} \biggl\{\frac{\bigl(\deg_{\wt{\Pbb},r}(x)\bigr)^2}{\vol_{\wt{\Pbb},r}(\mc{C})} \biggr\},
\end{equation}
where $\deg_{\wt{\Pbb},r}(x) := \int_{\Rd} \1\{\|y - x\| \leq r\} \,d\wt{\Pbb}(y)$ is the expected degree of $x$ in $\wt{G}_{n,r} = G_{n,r}[\mc{C}[X]]$.

\begin{proposition}
	\label{prop:sample_to_population}
	Fix $\delta \in (0,1)$. There exist universal constants $C_1$-$C_3$, and constants $b_1$,$b_2$ and $B_2$ that do not depend on $n$ or $\delta$, such that each of the following statements hold.
	\begin{itemize}
		\item With probability at least $1 - 3\exp\{-b_1\delta^2n\}$,
		\begin{equation}
		\label{eqn:sample_to_population_normalized_cut}
		\Phi_{n,r}(\mc{C}[X]) \leq (1 - C_1\delta) \Phi_{\Pbb,r}(\mc{C}[X]).
		\end{equation}
		\item With probability at least $1 - B_2n\exp\{-b_2\delta^2n\}$,
		\begin{equation}
		\label{eqn:sample_to_population_conductance}
		\Psi_{n,r}(\mc{C}[X]) \geq (1 - C_2\delta) \Psi_{\Pbb,r}(\mc{C}[X]),
		\end{equation}
		and 
		\begin{equation}
		\label{eqn:sample_to_population_local_spread}
		s_{n,r}(\mc{C}[X]) \geq (1 - C_3\delta) s_{\Pbb,r}(\mc{C}).
		\end{equation}
	\end{itemize}
\end{proposition}
A word on the proof techniques: the upper bound in~\eqref{eqn:sample_to_population_normalized_cut} follows by applying Hoeffding's inequality to control the deviations of $\cut_{n,r}(\mc{C}[X],\mc{C}^c[X])$, $\vol_{n,r}(\mc{C}[X])$ and $\vol_{n,r}(\mc{C}^c[X])$ around their expectations (noting that each of these is an order-$2$ U-statistic). To prove the lower bound~\eqref{eqn:sample_to_population_local_spread}, we require a union bound to control the minimum degree $\min_{u \in C} \{\deg(u;\wt{G}_{n,r})\}$, but otherwise the proof is similarly straightforward. On the other hand, the proof of~\eqref{eqn:sample_to_population_conductance} is considerably more complicated. We mention here only that our proof relies on the recent results of \citep{garciatrillos16b}, who upper bound the optimal transport distance between $\mathbb{P}_n$ and $\mathbb{P}$. For further details, we refer to Appendix~\textcolor{red}{(?)}, where we prove~\eqref{eqn:sample_to_population_conductance}, as well as~\citep{garciatrillos16}, who establish the asymptotic convergence of the sample conductance.

@@

\begin{enumerate}[label=(A\arabic*)]
	\item
	\label{asmp:bounded_density}
	\emph{Bounded density within cluster:} There exist constants
	$0<\lambda_{\sigma}< \Lambda_{\sigma}<\infty$ such that 
	$$
	\lambda_{\sigma} \leq \inf_{x \in \mc{C}_{\lambda,\sigma}} f(x) \leq \sup_{x \in \mc{C}_{\lambda,\sigma}} f(x)
	\leq \Lambda_{\sigma}.
	$$

>>>>

\subsection{Review: worst-case guarantees for local spectral clustering}
As mentioned, to date most analysis of local clustering has focused on worst-case guarantees, defined with respect to functionals of an a priori fixed graph $G = (V,E)$.  For instance, \cite{andersen2006} analyze the normalized cut of the cluster estimate $\wh{C}$ output by PPR. For a set $C \subseteq V$ with complement $C^c = V \!\setminus\! C$, the \emph{cut} and \emph{volume} are respectively,
\begin{equation}
\label{eqn:cut_volume}
\cut(C;G) := \sum_{u \in C} \sum_{v \in C^c}
\1\{(u,v) \in E\},~~ \vol(C; G) := \sum_{u \in C}  \sum_{v \in V} \1\{(u,v) \in E\},
\end{equation}
and the \emph{normalized cut} of $C$ is
\begin{equation}
\label{eqn:normalized_cut}
\Phi(C; G) := \frac{\cut(C;G)}{\min \set{\vol(C; G), \vol(C^c; G)}}.
\end{equation} 
\cite{andersen2006} show that when PPR is appropriately seeded within a candidate cluster $C \subseteq V$, the normalized cut of the candidate cluster is upper bounded $\Phi(\wh{C};G) \lesssim \sqrt{\Phi(C;G)}$. \cite{zhu2013} build on this: they introduce a second functional, the \emph{conductance} $\Psi(G)$, defined as 
\begin{equation}
\label{eqn:conductance}
\Psi(G) := \min_{S \subseteq V} \Phi(S;G),
\end{equation}
and show that if $\Phi(C;G)$ is much smaller than $\Psi(G[C])^2$---where $G[C] = (C,E \cap (C \times C))$ is the subgraph of $G$ induced by $C$--- then (in addition to having a small normalized cut) the cluster estimate $\wh{C}$ approximately recovers $C$. Our own analysis will build on that of~\cite{zhu2013}, and we will give a more detailed summary of their results in Section~\ref{sec:ub_symmetric_set_difference}.
For now, we merely reiterate that the conclusions of \citep{andersen2006,zhu2013} cannot be straightforwardly applied to our statistical setting, where the input data are random samples $\{x_1,\ldots,x_n\}$ drawn from a distribution $\mathbb{P}$, the graph $G$ is a random geometric graph formed by the user, and the candidate cluster is a set $\mc{C} \subset \Rd$.

@@

It is quite natural that $\Phi_{\Pbb,r}(\mc{C})$ and $\Psi_{\Pbb,r}(\mc{C})$ should help quantify the role geometry plays in local spectral clustering. Indeed, by construction these functionals are quite obviously the population-level analogues of the empirical quantities $\Phi_{n,r}(\mc{C}[X]) := \Phi(\mc{C}[X];G_{n,r})$ and $\Psi_{n,r}(\mc{C}[X]) = \Psi(G_{n,r}\bigl[\mc{C}[X]\bigr])$, and as we have already mentioned, these empirical quantities in turn suffice to upper bound the volume of the symmetric set difference. For this reason, similar population level functionals are used by \citep{shi2009,schiebinger2015,garciatrillos19} in the analysis of \emph{global} spectral clustering in a statistical context. We will comment more on the relationship between these works and our own results in~\ref{subsec:related_work}.

@@

@@

\subsection{PPR cluster recovery: the fixed graph case.}
\label{subsec:ppr_cluster_recovery_fixed_graph}
When PPR is run on a fixed graph $G = (V,E)$ with the goal of recovering a candidate cluster $C \subset V$, \cite{zhu2013} provide the sharpest known bounds on the volume of the symmetric set difference between the cluster estimate $\wh{C}$ and candidate cluster $C$. Since these results will play a major part in our analysis, in Lemma~\ref{lem:zhu} we restate them for the convenience of the reader.\footnote{Technically speaking, Lemma~3.4 of~\cite{zhu2013} differs from Lemma~\ref{lem:zhu} by some constant factors, and for completeness we prove Lemma~\ref{lem:zhu} in our appendix. Nevertheless, to be clear the essential idea of Lemma~\ref{lem:zhu} is no different than that of~\cite{zhu2013}, and we do not claim any novelty.}

In their most general form, the results of~\citet{zhu2013} depend on the mixing time of a lazy random walk over the induced subgraph $G[C]$. The \emph{mixing time} of a lazy random walk over a graph $G$ is
\begin{equation}
\label{eqn:mixing_time}
\tau_{\infty}(G) := \min\set{ t: \frac{{\pi}(u) - {q}_{v}^{(t)}(u)}
	{{\pi}(u)} \leq \frac{1}{4}, \; \text{for all $u,v \in V$}},
\end{equation}
where $q_v^{(t)} = e_v W^t$ is the distribution of a lazy random walk starting at node $v$ and running for $t$ steps, and $\pi = \lim_{t \to \infty} q_v^{(t)}$ is the limiting distribution of $q_v^{(t)}$. 
\begin{lemma}[Lemma~3.4 of \cite{zhu2013}]
	\label{lem:zhu}
	For a set $C \subseteq V$, suppose that
	\begin{equation}
	\label{eqn:zhu_condition}
	\alpha \leq \min\Bigl\{\frac{1}{45}, \frac{1}{2\tau_{\infty}(G[C])}\Bigr\},~~ \beta \leq \frac{1}{5\vol(C;G)}
	\end{equation}
	Then there exists a set $C^g \subset C$ with $\vol(C^g;G) \geq \frac{1}{2}\vol(C^g;G)$ such that for any $v \in C^g$, the sweep cut $S_{\beta,v}$ satisfies
	\begin{equation}
	\label{eqn:zhu_ub}
	\vol(S_{\beta,v} \vartriangle C;G) \leq 6\frac{\Phi(C;G)}{\alpha \beta}.
	\end{equation}
\end{lemma}
The upper bound in~\eqref{eqn:zhu_ub} does not obviously depend on the conductance $\Psi(G[C])$. However, as \cite{zhu2013} point out, letting $\pi_{\min}(G) := \min_{u \in V}\{\pi(u)\}$, it follows from Cheeger's inequality~\citep{chung1997} that 
\begin{equation}
\label{eqn:mixing_time_cheeger}
\tau_{\infty}(G) \leq \frac{\log(1/\pi_{\min}(G))}{\Psi(G)^2}.
\end{equation}
Therefore, setting (for instance) $\alpha = \frac{\Psi(G[C])^2}{2\log(1/\pi_{\min}(G))}$ and $\wh{C} = S_{\beta_0,v}$ for $\beta_0 = \frac{1}{5 \vol(C;G)}$, we obtain from~\eqref{eqn:zhu_ub} that 
\begin{equation}
\label{eqn:zhu_ub2}
\frac{\vol(C \vartriangle \wh{C};G)}{\vol(C; G)} \leq 60\frac{\Phi(C;G) \log\bigl( 1/\pi_{\min}(G[C])\bigr)}{\Psi(G[C])^2}.
\end{equation}

\paragraph{Improved bounds on mixing time.} Having reviewed the conclusions of~\cite{zhu2013}, we return now to our own setting, where the data is not a fixed graph $G$ but instead random samples $\{x_1,\ldots,x_n\}$, and our goal is to recover a candidate cluster $\mc{C} \subset \Rd$. Ideally, we would like to apply~\eqref{eqn:zhu_ub2} with $C = \mc{C}[X]$ and $G = G_{n,r}$, replace $\Phi_{n,r}(\mc{C}[X])$ and $\Psi_{n,r}(\mc{C}[X])$ by $\Phi_{\Pbb,r}(\mc{C})$ and $\Phi_{\Pbb,r}(\mc{C})$ inside~\eqref{eqn:zhu_ub2}, and thereby obtain an upper bound on $\Delta(\wh{C};\mc{C}[X])$ that depends only on $\Pbb$ and $\mc{C}$. Unfortunately, however, there is a catch: when the graph $G = G_{n,r}$ and the candidate cluster $C = \mc{C}[X]$, as $n \to \infty$ the sample normalized cut $\Phi_{n,r}(\mc{C}[X])$ and conductance $\Psi_{n,r}(\mc{C}[X])$ each converge to their population-level analogues, but $\pi_{\min}\bigl(G_{n,r}\bigl[\mc{C}[X]\bigr]\bigr) \lesssim 1/n$. Therefore the right hand side of~\eqref{eqn:zhu_ub2} grows at a rate of at least $\log n$, rendering~\eqref{eqn:zhu_ub2} a vacuous upper bound when the number of samples is sufficiently large.\footnote{To be clear: $\pi_{\min}(G)$ is obviously at most $1/|V|$ for any graph $G$ with $|V|$ total vertices, not merely $G_{n,r}\bigl[\mc{C}[X]\bigr]$. However, while in general it is certainly possible that $\Phi(C;G) \leq \frac{1}{\log(n)} \Psi(G[C])$, this will not be the case in our specific setting, where $G = G_{n,r}$ and $C = \mc{C}[X]$. Hence the need for a sharper bound.}

To fix this, we will improve the bound on mixing time given in~\eqref{eqn:mixing_time_cheeger}. While in the worst case (over an arbitrary input graph $G$) a ``start penalty'' of $\log(1/\pi_{\min}(G))$ is unavoidable, fortunately better bounds are possible when $G$ is a geometric graph such as $G_{n,r}$. Intuitively, this is because in geometric graphs, small sets are expanders---that is, if a set $R \subseteq V$ has sufficiently small volume, the normalized cut $\Phi(R;G)$ will be much larger than the conductance $\Psi(G)$. As a consequence, a random walk over $G$ will rapidly mix over all small sets $R$, and in our analysis of the mixing time we may therefore ``pretend'' as if the random walk were given a warm start over a larger set $S$. To delineate small sets $R$ from larger sets $S$, we introduce the \emph{local spread} functional $s(G)$, defined as
\begin{equation*}
s(G) := d_{\min}(G) \cdot \pi_{\min}(G),
\end{equation*}
where $d_{\min}(G) = \min_{u \in V}\bigl\{\deg(u;G)\bigr\}$, and likewise $d_{\max}(G) = \max_{u \in V}\bigl\{\deg(u;G)\bigr\}$.

Using a careful analysis we can make the intuition of the prior paragraph rigorous, leading to an improved upper bound on mixing time in Proposition~\ref{prop:pointwise_mixing_time}. The proof of this proposition, as with all results in this paper, is deferred to the appendix.
\begin{proposition}
	\label{prop:pointwise_mixing_time}
	Assume $d_{\max}(G)/d_{\min}(G)^2 \leq 1/16$. Then,
	\begin{equation}
	\label{eqn:pointwise_mixing_time}
	\tau_{\infty}(G) \leq \frac{8}{\Psi^2(G)} \ln \left(\frac{4}{s(G)}\right)\log_2 \left(\frac{8}{s(G)}\right)  + 4\log_2 \left(\frac{8}{s(G)}\right) + 1.
	\end{equation}
\end{proposition}

We emphasize that while Proposition~\ref{prop:pointwise_mixing_time} can be applied to \emph{any} graph $G$ (as long as the ratio of maximum degree to squared minimum degree is at most $1/16$), it is particularly useful when $G$ is a geometric graph. We give a precise upper bound on $s\bigl(G_{n,r}\bigl[\mc{C}[X]\bigr]\bigr)$ in Proposition~\ref{prop:sample_to_population_1}, which does not grow with $n$, and in combination with Proposition~\ref{prop:pointwise_mixing_time} this allows us to remove the unwanted $\log n$ factor from the upper bound in~\eqref{eqn:mixing_time}.

\subsection{Sample-to-population results}
\label{subsec:sample_to_population}
\textcolor{red}{(TODO): Fix $\deg_{\min}$ notation, and fix Propositions~\ref{prop:sample_to_population_1} and~\ref{prop:sample_to_population_2}.}\\

In Propositions~\ref{prop:sample_to_population_1} and~\ref{prop:sample_to_population_2}, we establish high probability bounds on the sample normalized cut, local spread, and conductance, in terms of their population-level analogues. We have already introduced the population normalized cut~\eqref{eqn:population_normalized_cut} and conductance~\eqref{eqn:population_conductance}; denoting $s_{n,r}(\mc{C}[X]) := s\bigl(G_{n,r}\bigl[\mc{C}[X]\bigr]\bigr)$, we now define the population local spread to be
\begin{equation}
\label{eqn:local_spread}
s_{\Pbb,r}(\mc{C}) := \min_{x \in \mc{C}} \biggl\{\frac{\bigl(\deg_{\wt{\Pbb},r}(x)\bigr)^2}{\vol_{\wt{\Pbb},r}(\mc{C})} \biggr\},
\end{equation}
where $\deg_{\wt{\Pbb},r}(x) := \int_{\Rd} \1\{\|y - x\| \leq r\} \,d\wt{\Pbb}(y)$ is the expected degree of $x$ in $G_{n,r}\bigl[\mc{C}[X]\bigr]$. In what follows, we use $b_1,b_2,\ldots$ and $B_1,B_2,\ldots$ to refer to constants that may depend on $\Pbb$ and $\mc{C}$, but do not depend on $n$ or $\delta$. We explicitly keep track of all constants in our proofs.
\begin{proposition}
	\label{prop:sample_to_population_1}
	Fix $\delta \in (0,1/3)$. There exist positive constants $b_1$-$b_3$ such that each of the following statements hold.
	\begin{itemize}
		\item With probability at least $1 - 3\exp\{-b_1\delta^2n\}$,
		\begin{equation}
		\label{eqn:sample_to_population_normalized_cut}
		\Phi_{n,r}(\mc{C}[X]) \leq (1 + 3\delta) \Phi_{\Pbb,r}(\mc{C}).
		\end{equation}
		\item With probability at least $1 - 2\exp\{-b_2\delta^2n\} - n\exp\{-b_3\delta^2n\}$,
		\begin{equation}
		\label{eqn:sample_to_population_local_spread}
		s_{n,r}(\mc{C}[X]) \geq (1 - 3\delta) s_{\Pbb,r}(\mc{C}).
		\end{equation}
	\end{itemize}
\end{proposition}
To lower bound the sample conductance $\Psi_{n,r}(\mc{C}[X])$, we impose the following regularity conditions on $\wt{\Pbb}$ and $\mc{C}$.
\begin{enumerate}[label=(A\arabic*)]
	\item 
	\label{asmp:domain} 
	The candidate cluster $\mc{C} \subseteq \Rd$ for $d \geq 2$, is a bounded, connected, open set with Lipschitz boundary. 
	\item 
	\label{asmp:bounded_density} 
	The distribution $\wt{\Pbb}$ has a density $\wt{f}: \mc{C} \to (0,\infty)$ such that there exist $f_{\min} \leq 1 \leq f_{\max}$ for which
	\begin{equation*}
	(\forall x \in \mc{C})~~ f_{\min} \leq \wt{f}(x) \leq f_{\max}.
	\end{equation*}
\end{enumerate}
Let $p_d := 3/4$ if $d = 2$, and else $p_d := 1/d$ for $d \geq 3$. 
\begin{proposition}
	\label{prop:sample_to_population_2}
	Fix $\delta \in (0,1/2)$. Suppose $\mc{C}$ and $\wt{\Pbb}$ satisfy~\ref{asmp:domain} and~\ref{asmp:bounded_density}. Then there exist positive constants $B_1$, $B_2$ that do not depend on $n$ or $\delta$, such that for any $n \in \mathbb{N}$ for which
	\begin{equation}
	\label{eqn:sample_to_population_conductance_sample_complexity}
	\frac{\log(n)^{p_d}}{n^{1/d}} \leq \delta \cdot B_1,
	\end{equation}
	the following inequality holds with probability at least $1 - B_2/n - (n + 2)\exp\{-b_2n/\sqrt{2}\}$:
	\begin{equation}
	\label{eqn:sample_to_population_conductance}
	\Psi_{n,r}(\mc{C}[X]) \geq (1 - 2\delta) \Psi_{\Pbb,r}(\mc{C}).
	\end{equation}
\end{proposition}

A word on the proof techniques: the upper bound in~\eqref{eqn:sample_to_population_normalized_cut} follows by applying Hoeffding's inequality to control the deviations of $\cut_{n,r}(\mc{C}[X],\mc{C}^c[X])$, $\vol_{n,r}(\mc{C}[X])$ and $\vol_{n,r}(\mc{C}^c[X])$ around their expectations (noting that each of these is an order-$2$ U-statistic). To prove the lower bound~\eqref{eqn:sample_to_population_local_spread}, we require a union bound to control the minimum degree $d_{\min}\bigl(G_{n,r}\bigl[\mc{C}[X]\bigr]\bigr)$, but otherwise the proof is similarly straightforward. On the other hand, the proof of~\eqref{eqn:sample_to_population_conductance} is considerably more complicated. We mention here only that our proof relies on the recent results of \citep{garciatrillos16b}, who upper bound the optimal transport distance between $\mathbb{P}_n$ and $\mathbb{P}$. For further details, we refer to Appendix~\textcolor{red}{(?)}, where we prove~\eqref{eqn:sample_to_population_conductance}, as well as~\citep{garciatrillos16}, who establish the asymptotic convergence of the sample conductance.

@@

\begin{enumerate}[label=(A\arabic*)]
	\item
	\label{asmp:lambda_bounded_density}
	\emph{Bounded density within cluster:} There exist constants
	$0<\lambda_{\sigma}< \Lambda_{\sigma}<\infty$ such that 
	$$
	\lambda_{\sigma} \leq \inf_{x \in \mc{C}_{\lambda,\sigma}} f(x) \leq \sup_{x \in \mc{C}_{\lambda,\sigma}} f(x)
	\leq \Lambda_{\sigma}.
	$$
